Conversational expectations account for apparent limits on theory of mind use

### 2017-08-26 writing/journal_manuscript/pnas-submission.tex
Reformat writing for pnas; revise abstract/sig statement

% arara: pdflatex
% arara: bibtex
% arara: pdflatex
% arara: pdflatex
% 
% Annual Cognitive Science Conference
% Sample LaTeX Paper -- Proceedings Format
%

% Original : Ashwin Ram (ashwin@cc.gatech.edu)       04/01/1994
% Modified : Johanna Moore (jmoore@cs.pitt.edu)      03/17/1995
% Modified : David Noelle (noelle@ucsd.edu)          03/15/1996
% Modified : Pat Langley (langley@cs.stanford.edu)   01/26/1997
% Latex2e corrections by Ramin Charles Nakisa        01/28/1997 
% Modified : Tina Eliassi-Rad (eliassi@cs.wisc.edu)  01/31/1998
% Modified : Trisha Yannuzzi (trisha@ircs.upenn.edu) 12/28/1999 (in process)
% Modified : Mary Ellen Foster (M.E.Foster@ed.ac.uk) 12/11/2000
% Modified : Ken Forbus                              01/23/2004
% Modified : Eli M. Silk (esilk@pitt.edu)            05/24/2005
% Modified : Niels Taatgen (taatgen@cmu.edu)         10/24/2006
% Modified : David Noelle (dnoelle@ucmerced.edu)     11/19/2014

%% Change "letterpaper" in the following line to "a4paper" if you must.

\documentclass[10pt,letterpaper]{article}

\usepackage{hyperref}
\usepackage{cogsci}
\usepackage{pslatex}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{apacite}
\usepackage{color}
\usepackage{todonotes}
\usepackage{dsfont}
\usepackage{array}
\usepackage{textcomp}
\usepackage{multirow}

\definecolor{Red}{RGB}{255,0,0}
\newcommand{\red}[1]{\textcolor{Red}{#1}}

\newcommand{\jd}[1]{\green{$^*$}\marginpar{\footnotesize{JD: \green{#1}}}}
\definecolor{Green}{RGB}{10,200,100}
\definecolor{Red}{RGB}{200,10,100}

\newcommand{\ndg}[1]{\textcolor{Green}{[ndg: #1]}}
\newcommand{\rdh}[1]{\textcolor{Red}{[rdh: #1]}}

\newcommand{\subsubsubsection}[1]{{\em #1}}
\newcommand{\eref}[1]{(\ref{#1})}
\newcommand{\tableref}[1]{Table \ref{#1}}
\newcommand{\figref}[1]{Figure \ref{#1}}
\newcommand{\appref}[1]{Appendix \ref{#1}}
\newcommand{\sectionref}[1]{Section \ref{#1}}

@@

\author{{\large \bf Robert X.~D.~Hawkins, Noah D.~Goodman}\\
  \{rxdh,ngoodman\}@stanford.edu\\
  Department of Psychology, 450 Serra Mall \\
  Stanford, CA 94305 USA}

@@

\maketitle

\begin{abstract}
Theory of mind is a powerful cognitive ability: by the age of six, people are capable of accurately reasoning about others' beliefs and desires. An influential series of language understanding experiments by Keysar and colleagues, however, showed that adults systematically failed to take a speaker's beliefs into account, revealing limitations on theory of mind. In this paper we argue that these apparent failures are in fact successes. Through a minimal pair of replications comparing scripted vs. unscripted speakers, we show that critical utterances used by Keysar and colleagues are uncooperative: they are less informative than what a speaker would actually produce in that situation. When we allow participants to naturally interact, we find that listener expectations are justified and errors are reduced. This ironically shows that apparent failures of theory of mind are in fact attributable to sophisticated expectations about speaker behavior---that is, to theory of mind.

\textbf{Keywords:} 
Theory of mind; social cognition; pragmatics
\end{abstract}

\section{Introduction}
\label{sec:intro}

\begin{figure*}[bt!]
\begin{center}
\includegraphics[scale = .37]{images/overall_view.png}
\end{center}
\vspace{-.5cm}
\caption{Interface used in the reported experiments. Objects behind the black squares were hidden from the director.}
\label{fig:interface}
\end{figure*}

@@

The argument offered by Keysar and colleagues is based on an elegant experimental paradigm, where participants played a simple communication game with a confederate. The two players were placed on opposite sides of a $4 \times 4$ grid containing a set of everyday objects (see Fig. \ref{fig:interface}). The confederate played the role of `director,' giving instructions about how to move objects around a grid, and the participant played the role of `matcher,' attempting to follow these instructions. For example, the objects in one trial included a cassette tape. The director gave an instruction like `move the tape up one square,' referring to the cassette.
Critically, some objects were occluded such that only the matcher could see them, creating an asymmetry in the players' knowledge. To perform accurately on critical trials, the matcher would need to apply theory of mind to reason about which objects were shared and which were private. 
For example, imagine a roll of tape were placed in an occluded slot: if a participant failed to account for the director's (partial) knowledge, she might interpret `tape' to mean the occluded roll of tape (which the director couldn't possibly know about). Indeed, \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} found that participants attempted to move the hidden item in 30\% of cases: 71\% of participants attempted to move this hidden item at least once (out of four critical cases) in the experimental condition, compared to 0\% in a control condition where there was no ambiguity over the referent. 
Additionally, eye-tracking data showed that participants considered the hidden item more often and for longer in the experimental condition than the control condition.

While these results are compelling, the paradigm has been criticized from few different angles. 
\citeA{HellerGrodnerTanenhaus08_Perspective} have pointed out that in many cases, the hidden object was a better fit for the referring expression than the one in common ground (e.g. the hidden roll of tape vs. the cassette tape for ``the tape''), making the hidden object \emph{a priori} more likely to be the referent; it would generally be fairer to compare two objects that fit the referring expression equally well. 
We validate this argument by empirically measuring relative fit of the expressions to the target and distractor items.
% Although relative fit was not empirically evaluated prior to our study,
Moreover, \citeA{HannaTanenhausTrueswell03_CommonGroundPerspective} argued that the viewpoint asymmetry paradigm is somewhat unnatural: common ground is typically built incrementally over the course of an interaction rather than presented all at once, and it is rare for a shared display to differ in perceptual accessibility. 
%Third, none of the experiments reported by Keysar and colleagues included a key comparison condition where the critical item (e.g. the roll of tape) was \emph{also} in common ground \cite{BrownSchmidtHanna11_IncrementalPerspectiveTaking}---perhaps people would refer to the critical item less in the privileged condition than in the full common ground condition, which would put a more positive spin on the results. %Note that none of these criticisms invalidate the paradoxical result that listeners move an occluded object, they just suggest an alternative explanation. Reference disambiguation is probabilistic: common ground and goodness of fit are weighed against one another.

In this paper, we offer an additional factor that helps account for Keysar's results. 
Theory of mind as applied within language understanding depends on an accurate model of what a speaker would say in different situations. Given an utterance, a listener can then reason backward to the most plausible situation \cite{Grice75_LogicConversation,Clark96_UsingLanguage,FrankGoodman12_PragmaticReasoningLanguageGames, GoodmanStuhlmuller13_KnowledgeImplicature}.
This suggests that we consider whether the utterances produced by the confederate in Keysar's critical conditions were actually what a speaker in that context would be expected to say. 
If not, then perhaps listeners are making choices that are in fact consistent with a correct pragmatic interpretation of the confederate's (uncooperative) utterance.
More precisely, when both players know that objects are occluded in the display, the speaker may tend to add additional precision to references in order to avoid confusion. If the listener \emph{expects} the speaker to do this, they will pragmatically pick the \emph{a priori} more likely referent of the referring expression, which in critical trials will be the occluded object. In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item. 
\begin{table*}
\begin{center}
\begin{tabular}{ p{2cm} | r | r |  r || r | r | r || r | r | r}
& \multicolumn{3}{c||}{\% attempted at least once} & \multicolumn{3}{c||}{\% attempted at least twice} & \multicolumn{3}{c}{\% of total cases}\\
\hline
& Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2  \\
Experimental & 71 & 93 & 61 & 46 & 57 & 32 & 30 & 43 & 24\\
Baseline        & 0   & 7   & 0   & 0   & 0   & 0   & 0   & 2   & 0\\
\end{tabular}
\caption{Side-by-side comparison of error rates in \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} and our two replications. The first two sections show the percentage of \emph{participants} attempting to move the occluded distractor at least once, or twice, of the four possible cases. The third section shows the percentage of \emph{all experimental trials} that the participant actually tried to move the occluded object. }
\label{table:mainResults}
\end{center}
\end{table*}

We began by replicating \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in a multi-player web experiment. We recruited participants to be both director and matcher (instead of using a confederate), but instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We replicated the original finding, but noted a tendency of directors to be overinformative in unscripted filler trials. We then ran the same experiment without using any scripted instructions, observing unconstrained director utterances. We found much greater precision in unconstrained director utterances, which match targets much better than distractors, and better performance of the matchers. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to sophisticated expectations about speaker behavior---that is, to theory of mind.

\section{Expt.~1: Scripted Replication}
\label{sec:Exp1}

\subsection{Participants}

@@

\subsection{Materials \& Procedures}

Participants interacted in a real-time, multi-player environment on the web \cite{Hawkins15_RealTimeWebExperiments}. 
%To deal with participants arriving at unpredictable times, we implemented a waiting room mechanism, where the first person to enter the game environment would see a screen saying ``Waiting for another player..." and the second person would trigger the game to start. 
Pairs of participants---assigned randomly to `director' and `matcher' roles---interacted with one another through a web interface, shown in Fig. \ref{fig:interface}. On the left side of the screen, participants could freely type messages to one another; on the right side the screen, players could view a set of objects placed in a 4 x 4 grid. Five of the grid cells were occluded from only the director's perspective, and the remaining 11 were visible to both matcher and director. Six or seven objects were displayed in the grid at a given time. One of these objects was a `target', such as a cassette tape, placed in an unoccluded cell such that both participants could see it. Another object, such as a roll of tape, was placed in an occluded slot such that it was only visible to the matcher. The rest of the objects were unrelated `fillers' placed in random locations. We used the same set of targets and occluded alternatives as \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse}, but we were unable to obtain the filler objects from the original experiment and created our own.
Before entering the game environment, every participant independently passed a short quiz about the task's instructions, ensuring that they understood the interface. Among other items on the quiz, we verified that both participants understood that items behind black cells were only visible to the matcher.

The experiment was composed of eight items, with each item using a different set of objects. Each item included one `critical pair' of objects, one of them the target and the other hidden, such as the cassette tape and the roll of tape. For each item, we gave the director a series of four instructions to move objects around, which were displayed as a series of arrows pointing from some object to an unoccupied cell. 
To collect clean mouse-tracking data, we began every instruction by asking the matcher to click a small circle in the center of the grid. After this small circle was clicked, the director was allowed to communicate the next instruction and we started recording from the matcher's mouse. 
One of the instructions was a `critical instruction,' which referred to the target object.
For half the instructions, directors were free to communicate however they wished. For the other half, including all the critical instructions, their messages to the matcher were pre-scripted using the precise wording from \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. For example, when giving instructions on how to move the cassette tape, the director would be forced to use the ambiguous utterance ``Move the tape down one square." (That is, in these conditions, the scripted message would automatically appear in the director's chat window, and they would have to click `send' for the experiment to continue.)

We collected baseline performance for each condition by replacing the hidden alternative (e.g.~a roll of tape) with an object that did not fit the critical instruction (e.g.~a battery); we used the same unambiguous replacements as \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. Each participant received half the items in the experimental condition and half in the baseline condition. The assignment of items to conditions was randomized across participants, and the order of conditions was randomized under the constraint that the same condition would not be used on more than two consecutive items. All object sets, object placements, and corresponding instruction sets were the same for all participants.

This paradigm differs from those used by \citeA{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in three primary ways. First, participants were not seated across from each other at a table: they each saw a view of the 4 x 4 grid on their screen and communicated via a text box. Second, we did not use a trained confederate. We randomly assigned one of the players to the role of the instruction giver, and maintained the original wording by scripting a subset of their instructions. Finally, the hidden object was not placed in a bag, in which respect our design more closely resembles \citeA{KeysarBarr___Brauner00_TakingPerspective}.

%The experiment proceeded through all eight items, with the director providing instructions and the participant moving the objects. Except for instructions with scripted messages, the participants were free to use the chat box for questions, small talk, or whatever else.
\begin{table*}
\begin{center}
\begin{tabular}{ m{1cm} | m{1.5cm} | m{1.5cm} | m{1.5cm} |  m{1.25cm} | m{1.6cm} | m{1.4cm} | m{1.1cm} | m{1.3cm} | m{1.3cm} |}
& & Item 1 & Item 2 & Item 3 & Item 4 & Item 5 & Item 6 & Item 7 & Item 8\\\hline
& instruction & ``glasses'' & ``bottom block'' & ``tape'' & ``large measuring cup'' & ``brush'' & ``eraser'' & ``small candle'' & ``mouse'' \\\hline
& target & sunglasses & block (3rd row) & cassette & medium cup & round hairbrush & board eraser & medium candle & computer mouse \\\hline
& hidden distractor & glasses case & block (4th row) & scotch-tape & large cup & flat \,\,\,hairbrush & pencil eraser & small candle & toy mouse \\
 \hline\hline
\multirow{2}{.75cm}{Expt.~1}  & \# incorrect & 0 & 5 & 1 & 3 & 2 &6 & 6 & 1 \\ 
                                               & \# correct & 6 & 3 & 1 & 8 & 2 & 2 & 4 & 6 \\
 \hline\hline
\multirow{2}{.75cm}{Expt.~2} & \# incorrect & 0 & 1 & 1 & 3 & 9 & 7 & 1 & 5 \\
                                              & \# correct   & 12&14&11&13&7 &8   &12& 8 \\

\end{tabular}
\vspace{-.5cm}
\caption{Item-wise error rates for critical trials}
\vspace{-.25cm}
\label{table:ItemWise}
\end{center}
\end{table*}

>>>>

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

\usepackage{pslatex}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{color}
\usepackage{todonotes}
\usepackage{dsfont}
\usepackage{array}
\usepackage{textcomp}
\usepackage{multirow}

@@

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Robert Hawkins}
\author[a]{Hyowon Gweon} 
\author[a]{Noah Goodman}

\affil[a]{Department of Psychology, Stanford University}

% Please give the surname of the lead author for the running footer
\leadauthor{Hawkins}

% TODO: Currently a bit too long (and probably not 'significant' sounding enough) 
\significancestatement{To communicate successfully, speakers must flexibly adjust the specificity of their utterances to the level demanded by the current context. In contexts where a speaker is unsure which objects their partner may be considering, it is helpful to provide additional disambiguating information: for example, saying "cassette tape" instead of the potentially ambiguous ``tape." %A listener who rationally expects such levels of specificity may, in turn, draw surprising inferences when they hear a less specific utterance than the situation demands. 
We argue that influential arguments for limits on Theory of Mind (ToM) in adults have neglected the nuanced pragmatics of the communication tasks they rest upon. A careful analysis of these pragmatics demands using a rational model of communication suggests a minimal pair of replications showing that apparent failures of visual perspective-taking are in fact successes for sophisticated social reasoning in context.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any conflict of interest here.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: author.two\@email.com}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Pragmatics $|$ Theory of mind $|$ Social cognition $|$ Replication}

\begin{abstract}
Humans are capable of accurately reasoning about other minds, but how ubiquitously do we use this ability in everyday life? Recent theories of pragmatic language use as recursive social reasoning remain in tension with evidence that adult listeners apparently fail to take their partner's perspective in communication. To reconcile these points of view, we focus attention on the problem faced by the \emph{speaker.} Through a minimal pair of replications comparing scripted vs. unscripted speakers, we show that scripted confederates used in previous studies are \emph{uncooperative}: they are less informative than what a speaker would actually produce in that situation. When we allow participants to naturally interact, we find that listener expectations are justified and errors are reduced. This ironically shows that apparent failures of theory of mind are in fact attributable to sophisticated expectations about speaker behavior--that is, to theory of mind. We formalize this argument in a computational model of pragmatic reasoning, and show that it provides a good fit with several recent findings.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

@@

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

@@

The argument offered by Keysar and colleagues is based on an elegant experimental paradigm, where participants played a simple communication game with a confederate. The two players were placed on opposite sides of a $4 \times 4$ grid containing a set of everyday objects (see Fig. \ref{fig:interface}). The confederate played the role of `director,' giving instructions about how to move objects around a grid, and the participant played the role of `matcher,' attempting to follow these instructions. For example, the objects in one trial included a cassette tape. The director gave an instruction like `move the tape up one square,' referring to the cassette.
Critically, some objects were occluded such that only the matcher could see them, creating an asymmetry in the players' knowledge. To perform accurately on critical trials, the matcher would need to apply theory of mind to reason about which objects were shared and which were private. 
For example, imagine a roll of tape were placed in an occluded slot: if a participant failed to account for the director's (partial) knowledge, she might interpret `tape' to mean the occluded roll of tape (which the director couldn't possibly know about). Indeed, \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} found that participants attempted to move the hidden item in 30\% of cases: 71\% of participants attempted to move this hidden item at least once (out of four critical cases) in the experimental condition, compared to 0\% in a control condition where there was no ambiguity over the referent. 
Additionally, eye-tracking data showed that participants considered the hidden item more often and for longer in the experimental condition than the control condition.

While these results are compelling, the paradigm has been criticized from few different angles. 
\cite{HellerGrodnerTanenhaus08_Perspective} have pointed out that in many cases, the hidden object was a better fit for the referring expression than the one in common ground (e.g. the hidden roll of tape vs. the cassette tape for ``the tape''), making the hidden object \emph{a priori} more likely to be the referent; it would generally be fairer to compare two objects that fit the referring expression equally well. 
We validate this argument by empirically measuring relative fit of the expressions to the target and distractor items.
% Although relative fit was not empirically evaluated prior to our study,
Moreover, \cite{HannaTanenhausTrueswell03_CommonGroundPerspective} argued that the viewpoint asymmetry paradigm is somewhat unnatural: common ground is typically built incrementally over the course of an interaction rather than presented all at once, and it is rare for a shared display to differ in perceptual accessibility. 
%Third, none of the experiments reported by Keysar and colleagues included a key comparison condition where the critical item (e.g. the roll of tape) was \emph{also} in common ground \cite{BrownSchmidtHanna11_IncrementalPerspectiveTaking}---perhaps people would refer to the critical item less in the privileged condition than in the full common ground condition, which would put a more positive spin on the results. %Note that none of these criticisms invalidate the paradoxical result that listeners move an occluded object, they just suggest an alternative explanation. Reference disambiguation is probabilistic: common ground and goodness of fit are weighed against one another.

In this paper, we offer an additional factor that helps account for Keysar's results. 
Theory of mind as applied within language understanding depends on an accurate model of what a speaker would say in different situations. Given an utterance, a listener can then reason backward to the most plausible situation \cite{Grice75_LogicConversation,Clark96_UsingLanguage,FrankGoodman12_PragmaticReasoningLanguageGames, GoodmanStuhlmuller13_KnowledgeImplicature}.
This suggests that we consider whether the utterances produced by the confederate in Keysar's critical conditions were actually what a speaker in that context would be expected to say. 
If not, then perhaps listeners are making choices that are in fact consistent with a correct pragmatic interpretation of the confederate's (uncooperative) utterance.
More precisely, when both players know that objects are occluded in the display, the speaker may tend to add additional precision to references in order to avoid confusion. If the listener \emph{expects} the speaker to do this, they will pragmatically pick the \emph{a priori} more likely referent of the referring expression, which in critical trials will be the occluded object. In other words, it is precisely \emph{because} the listener takes the speaker's mental state into consideration that they are tricked by an uncooperative confederate into choosing the wrong item. 
\begin{table*}
\begin{center}
\begin{tabular}{ p{2cm} | r | r |  r || r | r | r || r | r | r}
& \multicolumn{3}{c||}{\% attempted at least once} & \multicolumn{3}{c||}{\% attempted at least twice} & \multicolumn{3}{c}{\% of total cases}\\
\hline
& Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2 & Orig. & Expt. 1 & Expt. 2  \\
Experimental & 71 & 93 & 61 & 46 & 57 & 32 & 30 & 43 & 24\\
Baseline        & 0   & 7   & 0   & 0   & 0   & 0   & 0   & 2   & 0\\
\end{tabular}
\caption{Side-by-side comparison of error rates in \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} and our two replications. The first two sections show the percentage of \emph{participants} attempting to move the occluded distractor at least once, or twice, of the four possible cases. The third section shows the percentage of \emph{all experimental trials} that the participant actually tried to move the occluded object. }
\label{table:mainResults}
\end{center}
\end{table*}

We began by replicating \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in a multi-player web experiment. We recruited participants to be both director and matcher (instead of using a confederate), but instructions for critical items, as well as a random subset of filler items, remained scripted as in the original study. We replicated the original finding, but noted a tendency of directors to be overinformative in unscripted filler trials. We then ran the same experiment without using any scripted instructions, observing unconstrained director utterances. We found much greater precision in unconstrained director utterances, which match targets much better than distractors, and better performance of the matchers. This minimal pair of experiments demonstrates that listener mistakes are at least partially due to the pragmatics of the task, ironically showing that apparent failures of theory of mind are in fact attributable to sophisticated expectations about speaker behavior---that is, to theory of mind.

%% mention rubio fernandez early?
Using similar reasoning, \cite{RubioFernandez16_DirectorTaskAttention} showed that listeners became suspicious of what was actually in common ground after hearing overinformative utterances, thus successfully using theory of mind reasoning.

\section*{Referring in an uncertain visual environment}

In this section, I formalize the director-matcher task inside a rational speech act (RSA) model, where speakers produce informative utterances by reasoning about a rational listener agent who updates their beliefs about the world according to the literal semantics of the language. This allows us to test pragmatic accounts of task performance.

Previous theoretical advances in modeling pragmatics have typically hinged on `lifting' inference over some aspect of communication that was previously assumed to be fixed. For instance, \emph{manner implicatures} emerge from uncertainty over the true meanings of words in the lexicon \cite{BergenLevyGoodman16_LexicalUncertainty}, nonliteral constructions like \emph{hyperbole} and \emph{irony} emerge from uncertainty over the topic of discussion \cite{KaoWuBergenGoodman14_NonliteralNumberWords}, and \emph{prosody} and \emph{ellipsis} emerge from uncertainty (in the form of a noisy-channel model) over the various ways that a speaker's utterance could have been corrupted  \cite{BergenGoodman15_StrategicUseOfNoise}. Note that these are not ad hoc extensions; sources of uncertainty that are not strictly necessary to explain a particular effect are often omitted for simplicity, but they all coexist within the same model.

Pragmatic critiques of the director-matcher task point to an additional source of uncertainty that has not previously been explored: uncertainty over a communication partner's visual context. Considering this uncertainty, of course, \emph{is} theory of mind in action; our goal is to show how such a model built on social reasoning can produce the `failures' found by \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} as well as the phenomena observed by both \cite{HawkinsGoodman16_Keysar} and \cite{RubioFernandez16_DirectorTaskAttention}. For simplicity, we consider the full $2\times 2$ grid containing a single occluded cell, which was used by \cite{RubioFernandez16_DirectorTaskAttention}. We incorporate this source of uncertainty into both the speaker and listener models as follows:

\begin{enumerate}
\item A pragmatic speaker perceives an array of three objects $\mathcal{O}_S = \{o_i\}$, one of which is identified as a target $t$, but assumes that their partner sees an additional hidden object $o_h$ that is not in common ground. The speaker has uncertainty over the identity of this object, represented by a prior distribution $P(o_h)$ which they marginalize over when deciding what utterance to produce:
$$S_1(u | t, \mathcal{O}_p) \propto \sum_{o_h} L_0(t| \mathcal{O}_S \cup \{o_h\})P(o_h)P(u)$$

\item A pragmatic listener, reasoning about this uncertain speaker, is presented with a full array of four objects $\mathcal{O}_L$, one of which is identified as hidden to the speaker, $o_h$. The listener, however, also has some uncertainty over whether the experimenter is deceptive, represented by a Bernoulli random variable $d$, such that they assign a small prior probability $P(d)$ to the possibility that $o_h$ is actually in common ground. Upon hearing an utterance$u$, they update both their beliefs about which object is the target and whether the experimenter is being deceptive:
$$L_2(t, d | u, \mathcal{O}_L) = P(t)\left(S(u | t, \mathcal{O}_L)P(d)\right)^d(S(u | t, \mathcal{O}_L - o_h)(1-P(d))^{1-d}$$
\end{enumerate}
Technical details are omitted here for brevity, but the full model is fully specified in WebPPL and available to run at \url{http://forestdb.org/models/keysar.html}.

We focus on two critical patterns of results produced by the pragmatic listener $L_2$. First, we consider how it responds to the scripted instruction ``fish'' when presented with a context containing two fish (a red one that is occluded and a blue one in common ground) as well as two other unrelated objects. While it is most likely to choose the red fish in common ground, it has a relatively high probability (43\%) of choosing the hidden red fish, an effect \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} claimed was due to failure of theory of mind. This decision emerges in the model for two primary reasons. For one, a speaker with uncertainty over context could in principle be communicating about any object, and if they believed the hidden object was a red fish, the utterance `fish' would have a high probability of being produced. To a lesser extent, context uncertainty makes the overinformative utterance `blue fish' more likely than `fish', hence the omission of the color modifier has the (subtle) implication of \emph{not} the blue fish. This speaker phenomenon, while subtle under the parameter regime of the current implementation, produces the dynamic observed in the unscripted replication by  \cite{HawkinsGoodman16_Keysar}.

Second, we consider the listener's posterior over the \emph{deception} Bernoulli variable $d$. Just as \cite{RubioFernandez16_DirectorTaskAttention} found that listeners were more likely to suspect deception after hearing an `overinformative' utterance like `blue fish' when only one fish was in common ground, so too does our pragmatic listener. The pragmatic speaker is relatively more likely to produce the additional color modifier if they see multiple fish than when they see a single fish and simply have uncertainty over the unknown object. Hence, our model simultaneously captures all three phenomena.

\section*{Expt.~1: Scripted Replication}
\label{sec:Exp1}

\subsection*{Participants}

@@

\subsection*{Materials \& Procedures}

Participants interacted in a real-time, multi-player environment on the web \cite{Hawkins15_RealTimeWebExperiments}. 
%To deal with participants arriving at unpredictable times, we implemented a waiting room mechanism, where the first person to enter the game environment would see a screen saying ``Waiting for another player..." and the second person would trigger the game to start. 
Pairs of participants---assigned randomly to `director' and `matcher' roles---interacted with one another through a web interface, shown in Fig. \ref{fig:interface}. On the left side of the screen, participants could freely type messages to one another; on the right side the screen, players could view a set of objects placed in a 4 x 4 grid. Five of the grid cells were occluded from only the director's perspective, and the remaining 11 were visible to both matcher and director. Six or seven objects were displayed in the grid at a given time. One of these objects was a `target', such as a cassette tape, placed in an unoccluded cell such that both participants could see it. Another object, such as a roll of tape, was placed in an occluded slot such that it was only visible to the matcher. The rest of the objects were unrelated `fillers' placed in random locations. We used the same set of targets and occluded alternatives as \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}, but we were unable to obtain the filler objects from the original experiment and created our own.
Before entering the game environment, every participant independently passed a short quiz about the task's instructions, ensuring that they understood the interface. Among other items on the quiz, we verified that both participants understood that items behind black cells were only visible to the matcher.

The experiment was composed of eight items, with each item using a different set of objects. Each item included one `critical pair' of objects, one of them the target and the other hidden, such as the cassette tape and the roll of tape. For each item, we gave the director a series of four instructions to move objects around, which were displayed as a series of arrows pointing from some object to an unoccupied cell. 
To collect clean mouse-tracking data, we began every instruction by asking the matcher to click a small circle in the center of the grid. After this small circle was clicked, the director was allowed to communicate the next instruction and we started recording from the matcher's mouse. 
One of the instructions was a `critical instruction,' which referred to the target object.
For half the instructions, directors were free to communicate however they wished. For the other half, including all the critical instructions, their messages to the matcher were pre-scripted using the precise wording from \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. For example, when giving instructions on how to move the cassette tape, the director would be forced to use the ambiguous utterance ``Move the tape down one square." (That is, in these conditions, the scripted message would automatically appear in the director's chat window, and they would have to click `send' for the experiment to continue.)

We collected baseline performance for each condition by replacing the hidden alternative (e.g.~a roll of tape) with an object that did not fit the critical instruction (e.g.~a battery); we used the same unambiguous replacements as \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}. Each participant received half the items in the experimental condition and half in the baseline condition. The assignment of items to conditions was randomized across participants, and the order of conditions was randomized under the constraint that the same condition would not be used on more than two consecutive items. All object sets, object placements, and corresponding instruction sets were the same for all participants.

This paradigm differs from those used by \cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse} in three primary ways. First, participants were not seated across from each other at a table: they each saw a view of the 4 x 4 grid on their screen and communicated via a text box. Second, we did not use a trained confederate. We randomly assigned one of the players to the role of the instruction giver, and maintained the original wording by scripting a subset of their instructions. Finally, the hidden object was not placed in a bag, in which respect our design more closely resembles \cite{KeysarBarr___Brauner00_TakingPerspective}.

%The experiment proceeded through all eight items, with the director providing instructions and the participant moving the objects. Except for instructions with scripted messages, the participants were free to use the chat box for questions, small talk, or whatever else.
% \begin{table*}
% \begin{center}
% \begin{tabular}{ m{1cm} | m{1.5cm} | m{1.5cm} | m{1.5cm} |  m{1.25cm} | m{1.6cm} | m{1.4cm} | m{1.1cm} | m{1.3cm} | m{1.3cm} |}
% & & Item 1 & Item 2 & Item 3 & Item 4 & Item 5 & Item 6 & Item 7 & Item 8\\\hline
% & instruction & ``glasses'' & ``bottom block'' & ``tape'' & ``large measuring cup'' & ``brush'' & ``eraser'' & ``small candle'' & ``mouse'' \\\hline
% & target & sunglasses & block (3rd row) & cassette & medium cup & round hairbrush & board eraser & medium candle & computer mouse \\\hline
% & hidden distractor & glasses case & block (4th row) & scotch-tape & large cup & flat \,\,\,hairbrush & pencil eraser & small candle & toy mouse \\
%  \hline\hline
% \multirow{2}{.75cm}{Expt.~1}  & \# incorrect & 0 & 5 & 1 & 3 & 2 &6 & 6 & 1 \\ 
%                                                & \# correct & 6 & 3 & 1 & 8 & 2 & 2 & 4 & 6 \\
%  \hline\hline
% \multirow{2}{.75cm}{Expt.~2} & \# incorrect & 0 & 1 & 1 & 3 & 9 & 7 & 1 & 5 \\
%                                               & \# correct   & 12&14&11&13&7 &8   &12& 8 \\

% \end{tabular}
% \vspace{-.5cm}
% \caption{Item-wise error rates for critical trials}
% \vspace{-.25cm}
% \label{table:ItemWise}
% \end{center}
% \end{table*}

\subsection*{Results and discussion}
In Table \ref{table:mainResults} we show the error rates on critical items, and compare to the data from
\cite{KeysarLinBarr03_LimitsOnTheoryOfMindUse}.
We find that 93\% of participants (all but one) attempted to move the hidden distractor at least once in the Experimental condition, out of four possible items, compared to only 7\% (only one) in the baseline condition. This is similar to the effect observed by the authors in the original study, which found 71\% and 0\%. Our errors were larger across the board, perhaps due to the interface or the population, but the gap between the two conditions is roughly the same size.
%Next, we examine whether there are differences in error rates across items. 
In Table \ref{table:ItemWise}, we break down the pattern of errors by item. 
%Because our randomization procedure undersampled the experimental condition for several items, a chi-squared test is inappropriate. Using Fisher's exact test, we cannot reject the null hypothesis of independence ($p = 0.13$). Still, 
We note that several items have much higher error rates than others -- for example, 75\% of participants in the experimental condition of item 6 made an error (the ``whiteboard eraser'' vs. the ``pencil eraser'') while only 17\% of participants in item 8 made an error (the ``computer mouse'' vs. the ``toy mouse''). Informally, it seems as though the more difficult items are the ones where the utterance fits the distractor better than the target. We empirically substantiate this observation in our results for Expt.~2 below.
